---
title: "Bioinformatics Workshop: Working with FASTA Files in R"
date: '2024-07-12'
---

Introduction to FASTA Files and Basic Analysis
Objective:
- Introduce basic bioinformatics techniques for handling FASTA files using R
- Downloading, reading, and performing basic analysis on FASTA files
- Explore sequence extraction, pattern matching, and sequence manipulation
- Use visualization techniques to understand sequence compositions


Load necessary libraries
``````{r}
# Install and load necessary libraries
if (!require("Biostrings")) install.packages("Biostrings")
if (!require("seqinr")) install.packages("seqinr")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("dplyr")) install.packages("dplyr")
if (!require("rentrez")) install.packages("rentrez")
if (!require("tibble")) install.packages("tibble")
if (!require("tidyr")) install.packages("tidyr")
if (!require("seqinr")) install.packages("seqinr")
if (!requireNamespace("IRanges", quietly = TRUE)) {
  BiocManager::install("IRanges")
}
if (!requireNamespace("annotate", quietly = TRUE)) {
  BiocManager::install("annotate")
}

library(annotate)
library(Biostrings)
library(seqinr)
library(ggplot2)
library(dplyr)
library(rentrez)
library(tibble)
library(tidyr)
library(seqinr)
library(IRanges)
```

Download and Read a Sample FASTA File from NCBI
```{r}
# Download a FASTA file from NCBI
ncbi_id <- "NC_012920"  # Example: human mitochondrial DNA
fasta_data <- entrez_fetch(db="nuccore", id=ncbi_id, rettype="fasta")
write(fasta_data, file = "sample_NC_012920.fasta")

# Read the FASTA file into R
fasta_data <- readDNAStringSet("sample_NC_012920.fasta")
print(fasta_data)
```

Download a FASTA file with multiple sequences from NCBI and save in same file 
```{r}
# Define the NCBI IDs for the sequences you want to retrieve
ncbi_ids <- c("NM_001200001", "NM_001354762", "NM_001126112")  # Example NCBI IDs, replace with your desired IDs
# NM_001200001: mRNA sequence of the human beta-globin gene (HBB)
# NM_001354762: mRNA sequence of the human hemoglobin subunit delta (HBD) gene
# NM_001126112: mRNA sequence of the human myoglobin gene (MB)

# Fetch the sequence data in FASTA format from the Nucleotide Core database
multiple_fasta_data <- entrez_fetch(db = "nuccore", id = ncbi_ids, rettype = "fasta")

# Print the retrieved FASTA data
cat(multiple_fasta_data)

# Save the FASTA data to a file
fasta_file <- "multiple_sequences.fasta"
write(multiple_fasta_data, file = fasta_file)

# Confirm the file has been saved
cat("FASTA file saved as:", fasta_file)

```

Download and Read a Sample FASTA File from other avaibale databases
```{r}
# Choose the Swiss-Prot database
choosebank("swissprot")

# Define the accession number for the hemoglobin sequence
accession_number <- "P69891"

# Query the Swiss-Prot database for the hemoglobin sequence
hemo <- query("hemoglo", paste("AC=", accession_number, sep=""))

# Check the structure of the query result
str(hemo)

# To get the name, length, and other attributes for every available sequence in the query
attributes(hemo$req)

# To get the sequence of the protein
seq_hemo <- getSequence(hemo$req[[1]])
print(seq_hemo)

# To get annotation
annots <- getAnnot(hemo$req[[1]])
print(annots)

# Close the database to avoid multiple open banks and for faster data processing
closebank()
```

Basic Sequence Analysis
Count the number of sequences and calculate the length for a FASTA with single sequence 
```{r}
# Count the number of sequences
num_sequences <- length(fasta_data)
print(paste("Number of sequences:", num_sequences))

# Calculate the length of each sequence
sequence_lengths <- width(fasta_data)
print(sequence_lengths)
```

Count the number of sequences and calculate the length for a FASTA file with multple sequences
```{r}
# Split the fetched data into individual sequences using the FASTA header lines as delimiters
sequences <- strsplit(multiple_fasta_data, split = ">")[[1]]
sequences <- sequences[sequences != ""]  # Remove empty elements

# Function to extract sequence length
get_sequence_length <- function(fasta_sequence) {
  # Split the sequence into lines and remove the first line (header)
  lines <- unlist(strsplit(fasta_sequence, "\n"))
  sequence <- paste(lines[-1], collapse = "")
  return(nchar(sequence))
}

# Calculate the lengths of the sequences
sequence_lengths <- sapply(sequences, get_sequence_length)

# Count the number of sequences
num_sequences <- length(sequence_lengths)

# Print the results
cat("Number of sequences:", num_sequences, "\n")
cat("Lengths of sequences:", sequence_lengths, "\n")
```

Find a Specific Sequence
Search for a specific sequence within the FASTA data.
```{r}
# Search for a specific sequence
search_sequence <- "ACT"
matches <- vmatchPattern(search_sequence, fasta_data)
print(matches)

# Search for a specific sequence
search_sequence <- "TCGTT"
matches <- vmatchPattern(search_sequence, multiple_fasta_data)
print(matches)
```

Search multiple sequences in a fasta file and get their coordinates
```{r}
# Load the sequences into a DNAStringSet object
dna_sequences <- readDNAStringSet("multiple_sequences.fasta", format = "fasta")

# Search for specific sequences and store results with NCBI IDs
search_sequences <- c("ACT", "TCGTT")  # Add as many motifs of interest 
results <- list()

for (search_sequence in search_sequences) {
  matches <- vmatchPattern(DNAString(search_sequence), dna_sequences)
  
  # Extract NCBI IDs of matching sequences
  full_names <- names(dna_sequences)[lengths(matches) > 0]
  matching_ids <- sapply(strsplit(full_names, " "), `[`, 1)
  
  # Extract start and stop coordinates
  start_coords <- lapply(matches[lengths(matches) > 0], start)
  end_coords <- lapply(matches[lengths(matches) > 0], end)
  
  # Store the result
  result <- data.frame(
    NCBI_ID = rep(matching_ids, lengths(matches[lengths(matches) > 0])),
    Start = unlist(start_coords),
    End = unlist(end_coords),
    Sequence_Match = search_sequence
  )
  
  results[[search_sequence]] <- result
  
  # Print the results
  cat("Results for sequence:", search_sequence, "\n")
  print(result)
}

# Combine all results into a single data frame
combined_results <- do.call(rbind, results)

# Print the final combined results
cat("Combined Results:\n")
print(combined_results)

# Save the combined results to a CSV file
csv_file <- "sequence_matches.csv"
write.csv(combined_results, file = csv_file, row.names = FALSE)

cat("Results saved to:", csv_file, "\n")
```

Extract and Isolate Sequences
Extract specific sequences
```{r}
# Extract a specific sequence by index
specific_sequence <- dna_sequences[[1]] #  extract elements from a list
cat("Specific sequence:\n", as.character(specific_sequence), "\n") # DNAString object to a character string
cat("Number of nucleotides in specific sequence:", nchar(as.character(specific_sequence)), "\n")
```

Isolate Sequences
```{r}
dummy_fasta <- "
>seq1
TGCGTACGTAGCTAGCTAGCTAGCTAGCTGACTGACTGACTGAC
>seq2
ATGCGTACGTAGCTAGCTAGCTAGCTAGCTGACTGACTGACTGACGATGCGTACGTAGCTAGCTAGCTAGCTAGCTGACTGACTGACTGAC
>seq3
ATGCGTACGTAGCTAGCTAGCTAGCTAGCTGACTGACTGACTGACGATGCGTACGTAGCTAGCTAGCTAGCTAGCTGACTGACTGACTGACGATGCGTACGTAGCTAGCTAGCTAGCTAGCTGACTGACTGACTGAC
>seq4
TGCGTACGTAGCTAGCTAGCTAGCTAGCTGACTGACTGACTGACGATGCGTACGTAGCTAGCTAGCTAGCTAGCTGACTGACTGACTGACGATGCGTACGTAGCTAGCTAGCTAGCTAGCTGACTGACTGACTGACGATGCGTACGTAGCTAGCTAGCTAGCTAGCTGACTGACTGACTGACGATGCGTACGTAGCTAGCTAGCTAGCTAGCTGACTGACTGACTGAC"
writeLines(dummy_fasta, con = "dummy_sequences.fasta")
# 44, 88, 132, 176 sequence lengths

# Load the dummy sequences into a DNAStringSet object
dna_sequences <- readDNAStringSet("dummy_sequences.fasta", format = "fasta")

# Isolate sequences longer than a specified length (e.g., 100 bp)
long_sequences <- dna_sequences[width(dna_sequences) > 100]
cat("Long sequences (more than 100 bp):\n")
for (i in seq_along(long_sequences)) {
  cat("Sequence", i, ":\n", as.character(long_sequences[[i]]), "\n")
  cat("Number of nucleotides:", nchar(as.character(long_sequences[[i]])), "\n")
}
```

Save Manipulated FASTA File
Save the extracted or manipulated sequences to a new FASTA file.
```{r}
# Save the long sequences to a new FASTA file
writeXStringSet(long_sequences, filepath = "long_sequences.fasta")
```

Load and Read a Real Dataset: Human Mitochondrial DNA (mtDNA)
```{r}
# Download the human mitochondrial DNA (mtDNA) FASTA file from NCBI
ncbi_id <- "NC_012920"  # Human mitochondrial DNA
fasta_data <- entrez_fetch(db="nuccore", id=ncbi_id, rettype="fasta")
write(fasta_data, file = "human_mtDNA.fasta")

# Read the mtDNA FASTA file
mtDNA <- readDNAStringSet("human_mtDNA.fasta")
print(mtDNA)
```

AT and GC Content Analysis
Calculate the AT and GC content of the mtDNA sequences.
```{r}
# Calculate GC and AT content
gc_content <- letterFrequency(mtDNA, letters = c("G", "C"), as.prob = TRUE)
at_content <- letterFrequency(mtDNA, letters = c("A", "T"), as.prob = TRUE)

# Combine GC and AT content into one data frame
content_df <- as.data.frame(cbind(gc_content, at_content))
colnames(content_df) <- c("G", "C", "A", "T")

# Add GC_Content and AT_Content columns
content_df <- content_df %>%
  rownames_to_column("Sequence") %>%
  mutate(GC_Content = G + C, AT_Content = A + T)

# Print the content data frame
print(content_df)
```

Nucleotide Content Analysis
Calculate the nucleotide content of the mtDNA sequences.
```{r}
# Reshape the data for plotting proportions of A, T, G, and C
nucleotide_df <- content_df %>%
  dplyr::select(Sequence, G, C, A, T) %>%
  pivot_longer(cols = c(G, C, A, T), names_to = "Nucleotide", values_to = "Proportion")

# Visualize proportions of A, T, G, and C
ggplot(nucleotide_df, aes(x = Sequence, y = Proportion, fill = Nucleotide)) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Nucleotide Proportions of mtDNA Sequences") +
  xlab("Sequence Index") +
  ylab("Proportion (%)") +
  theme_minimal() +
  scale_fill_manual(values = c("A" = "green", "T" = "yellow", "G" = "blue", "C" = "red"))
```

Find a Motif
Search for a specific motif within the mtDNA sequences.
```{r}
# Search for a motif
motif <- "ATG"
motif_matches <- vmatchPattern(motif, mtDNA)
print(motif_matches)

motif <- "ATTGCG"
motif_matches <- vmatchPattern(motif, mtDNA)
print(motif_matches)
```

Additional Analysis and Visualization

Nucleotide Frequency
Calculate the frequency of each nucleotide in the sequences.
```{r}
# Calculate nucleotide frequency
nucleotide_freq <- alphabetFrequency(mtDNA)
print(nucleotide_freq)
```

Sequence Composition
Visualize the composition of sequences in terms of nucleotide percentages.
```{r}
# Calculate nucleotide composition
composition <- letterFrequency(mtDNA, letters = c("A", "T", "G", "C"), as.prob = TRUE)
composition_df <- as.data.frame(composition)
composition_df <- composition_df %>% 
  rownames_to_column("Sequence") %>% 
  pivot_longer(cols = -Sequence, names_to = "Nucleotide", values_to = "Percentage")

# Visualize nucleotide composition
ggplot(composition_df, aes(x = Sequence, y = Percentage, fill = Nucleotide)) +
  geom_bar(stat = "identity", position = "fill") +
  ggtitle("Nucleotide Composition in mtDNA Sequences") +
  xlab("Sequence Index") +
  ylab("Percentage") +
  theme_minimal()
```

Sequence Pattern Matching
Identify and visualize the occurrence of a specific pattern within the sequences.
```{r}
# Identify pattern matches
pattern <- "TTG"
pattern_matches <- vmatchPattern(pattern, mtDNA)
print(pattern_matches)

# Visualize pattern matches
pattern_count <- sapply(pattern_matches, length)
pattern_count_df <- data.frame(Sequence = names(pattern_count), Count = pattern_count)

ggplot(pattern_count_df, aes(x = Sequence, y = Count, fill = Sequence)) +
  geom_bar(stat = "identity") +
  ggtitle("Pattern 'TTG' Occurrences in mtDNA Sequences") +
  xlab("Sequence Index") +
  ylab("Pattern Count") +
  theme_minimal()
```
```{r}
# Define patterns to search for
patterns <- c("TTG", "GAC", "CTG")

# Initialize a list to store pattern counts
pattern_counts_list <- list()

# Identify pattern matches and count occurrences
for (pattern in patterns) {
  pattern_matches <- vmatchPattern(pattern, mtDNA)
  pattern_count <- sapply(pattern_matches, length)
  pattern_counts_list[[pattern]] <- pattern_count
}

# Combine pattern counts into a single data frame
pattern_count_df <- do.call(rbind, lapply(names(pattern_counts_list), function(pattern) {
  data.frame(Sequence = names(pattern_counts_list[[pattern]]), Count = pattern_counts_list[[pattern]], Pattern = pattern)
}))

# Convert Sequence to a factor to maintain order in the plot
pattern_count_df$Sequence <- factor(pattern_count_df$Sequence, levels = names(pattern_count))

# Visualize pattern matches
ggplot(pattern_count_df, aes(x = Sequence, y = Count, fill = Pattern)) +
  geom_bar(stat = "identity", position = "dodge") +
  ggtitle("Pattern Occurrences in mtDNA Sequences") +
  xlab("Sequence Index") +
  ylab("Pattern Count") +
  theme_minimal()
```

----------------------------------------- Exercise -----------------------------------------

Dr. Chris is investigating genetic markers in mitochondrial DNA. He needs your help to analyze the mtDNA data and identify specific patterns.

a. GC Content Calculation
Dr. Chris wants to analyze the GC content across different sequences. Calculate the GC content for each sequence in the mtDNA dataset. Identify sequences with unusually high or low GC content.

```{r}
# Solution a




# Hint: Look for sequences with GC content significantly different from the average, indicating potential areas of interest.
```

b. Motif Search
Dr. Chris suspects that a specific motif is prevalent in the mtDNA sequences. Search for the motif "ATG" and identify its locations within the sequences.
```{r}
# Solution b




# Hint: Notice any sequences with a high frequency of the motif, indicating regions of interest.
```

c. Calculate AT/GC Ratios and Nucleotide Composition

Dr. Chris wants to compare AT and GC content ratios and visualize the nucleotide composition. Calculate the AT/GC ratios for each sequence in the mtDNA dataset and create a stacked bar plot to show the AT/GC content percentages. Also, create individual bar plots for each nucleotide.
```{r}
# Solution c




# Hint: Use dplyr::select to avoid conflicts with other packages and ensure correct column selection. Calculate nucleotide percentages for plotting by dividing each nucleotide count by the total count.
```


d. Identify Palindromic Sequences
Dr. Chris is interested in palindromic sequences within the mtDNA. Identify and list all palindromic sequences.
Dummy data
```{r}
# Create a dummy DNAStringSet with known palindromic and non-palindromic sequences
dummy_sequences <- DNAStringSet(c(
  "ATGCAT",         # Palindromic sequence
  "CGTAGCTACG",     # Palindromic sequence
  "AGCTCGTCA",      # Non-palindromic sequence
  "AATTCCGGAATT",   # Palindromic sequence
  "GATTACA"         # Non-palindromic sequence
))
```

```{r}
# Solution d



# Hint: Palindromic sequences read the same forwards and backwards when considering the reverse complement. Use reverseComplement to check for palindromic sequences in the dataset.
```

