---
title: "Bioinformatics Visualization in R - Part 2"
date: '2024-07-12'
---

Lecture 2: 
Introduction to Visualization

Objective: 
 - Explore advanced visualization techniques and create more complex bioinformatics plots using biological data.
   Density Plot, Violin Plot, Dendogram, Heatmap
 - Overview of syntax with simulated data
 - Explore plots using iris dataset.
 - Exercise 2

Load necessary libraries
```{r}
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("GGally")) install.packages("GGally")
if (!require("pheatmap")) install.packages("pheatmap")
if (!require("dendextend")) install.packages("dendextend")
if (!require("ggridges")) install.packages("ggridges")

library(dendextend)
library(ggplot2)
library(GGally)
library(pheatmap)
library(reshape2)
library(dplyr)
```

Generate the Mock Gene Expression Dataset
```{r}
# Set random seed for reproducibility
set.seed(42)

# Number of genes and samples
num_genes <- 10
num_samples <- 5

# Generate gene names and sample names
genes <- paste0("Gene_", 1:num_genes)
samples <- paste0("Sample_", 1:num_samples)

# Generate random expression levels
expression_data <- matrix(rnorm(num_genes * num_samples), nrow = num_genes, ncol = num_samples)
colnames(expression_data) <- samples
rownames(expression_data) <- genes

# Convert to data frame
df_expression <- as.data.frame(expression_data)

# Display the first few rows of the dataset
head(df_expression)

```

Visualization 
Part 2: Advanced Visualization Techniques

5. Density Plot
   A density plot shows how the data is spread out, highlighting where values are concentrated.
```{r }
df_melted <- melt(df_expression, variable.name = "Sample", value.name = "Expression")

# Create the density plot
ggplot(df_melted, aes(x = Expression)) +
  geom_density(fill = "blue", alpha = 0.5) + # transparency
  labs(title = "Density Plot of Gene Expression Levels",
       x = "Expression Level",
       y = "Density") +
  theme_minimal()
```

6. Violin Plot:
   A violin plot is a data visualization that combines a box plot and a kernel density plot to show the distribution, probability density, and variability of data across different categories.
```{r }
# Melt the data frame for ggplot2
df_melted <- melt(df_expression, variable.name = "Sample", value.name = "Expression")

# Create the violin plot with colors
ggplot(df_melted, aes(x = Sample, y = Expression, fill = Sample)) +
  geom_violin(trim = FALSE) +
  labs(title = "Violin Plot of Gene Expression Levels",
       x = "Sample",
       y = "Expression Level") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3")
```


7. Dendrogram: 
   A dendrogram is a tree-like diagram that displays the arrangement of clusters formed by hierarchical clustering, 
   showing the relationships and distances between data points.
```{r }
# Generate the hierarchical clustering
hc <- hclust(dist(df_expression), method = "ward.D2")

# Create the dendrogram
dend <- as.dendrogram(hc)
plot(dend, main = "Dendrogram", xlab = "Genes", ylab = "Distance")
```


8. Heatmap + Dendrogram
   Heatmaps are useful to visualize matrix-like data, such as gene expression data
```{r }
# Create the heatmap
pheatmap(df_expression, scale = "row", main = "Heatmap of Gene Expression")
```


# Load the iris dataset and prepare it for visualization:

```{r }
# Load the iris dataset
data("iris")

# Display the first few rows of the dataset
head(iris)
summary(iris)
```
5. Density Plot
```{r }
# Melt the data frame for ggplot2
df_melted <- melt(iris, id.vars = "Species", variable.name = "Measurement", value.name = "Value")

# Create the density plot with facets for each species
ggplot(df_melted, aes(x = Value, fill = Measurement)) +
  geom_density(alpha = 0.5) + # transparency
  labs(title = "Density Plot of Iris Measurements by Species",
       x = "Value",
       y = "Density") +
  theme_minimal() +
  facet_wrap(~ Species)
```


6. Density Plot
```{r }
# Melt the data frame for ggplot2
df_melted <- melt(iris, id.vars = "Species", variable.name = "Measurement", value.name = "Value")

# Create the violin plot with colors
ggplot(df_melted, aes(x = Measurement, y = Value, fill = Species)) +
  geom_violin(trim = FALSE) +
  labs(title = "Violin Plot of Iris Measurements",
       x = "Measurement",
       y = "Value") +
  theme_minimal() +
  facet_wrap(~ Measurement, scales = "free")
```


7. Dendogram
```{r }
# Remove the Species column for clustering
iris_no_species <- iris[, -5]

# Compute the correlation matrix
cor_matrix <- cor(iris_no_species)
cor_matrix

# Perform hierarchical clustering on the correlation matrix
hc <- hclust(as.dist(1 - cor_matrix), method = "ward.D2")

# Create the dendrogram
dend <- as.dendrogram(hc)

# Plot the dendrogram
plot(dend, main = "Dendrogram of Iris Attributes", xlab = "Attributes", ylab = "Distance")
```



```{r}

# Remove the Species column for heatmap plotting
iris_data <- iris[, -5]

# Create the heatmap with a simplified color scale
pheatmap(as.matrix(iris_data), 
         main = "Heatmap of Iris Measurements", 
         cluster_rows = TRUE, cluster_cols = TRUE, 
         show_rownames = FALSE,
         color = colorRampPalette(c("blue", "white", "red"))(50))
```


---------------------------------------Assignment 2: Part 2--------------------------------------------------------

Dr. Smith is studying the famous iris dataset to understand the differences in flower measurements across three species: setosa, versicolor, and virginica. Help Dr. Smith visualize and interpret the data to identify distinguishing features of each species.

a. Density Plot Analysis:
   Create a density plot to visualize the distribution of sepal length measurements across the three iris species. 
   Identify which species have similar or distinct sepal length distributions.

```{r }
# Solution a


# Hint: Look for peaks in the density plots to see which species have similar or distinct sepal length distributions.
```


b. Violin Plot Analysis:
   Use a violin plot to compare the distribution and density of petal widths across the three iris species. 
   Highlight which species have the widest and narrowest petals.

```{r }
# Solution b


# Hint: Examine the width and shape of the violins to understand the distribution and density of petal widths across species.\
```

