---
title: "Script_3"
author: "Hamed Abdollahi"
date: "2024-07-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
x <- c("Apple", "Strawberry", "Cherry")

for (fruit in x)
  print(paste("I like", fruit))# without curly braces if the for loop is a single statement.
```

```{r}
for (i in LETTERS[1:10]) {
  print(i)
}
```

```{r}
x <- LETTERS[1:10]#Vector Definition
for(i in 5:9) {#For Loop
       ## Indexing 'x'
       print(x[i])  
}
```

The `seq_along()` function is quite useful in R for iterating over elements of an object.

It is use to generate a sequence of integers that correspond to the indices of an object.
So it is based on the length of the object.

```{r}
x <- c("A", "B", "C", "D", "E")

for (i in seq_along(x)) {
  print(paste("Index", i, ":", x[i]))
}
```

Usage in genomics

```{r}
# Load the iris dataset (if not already loaded)
data(iris)
str(iris)
library(dplyr)
glimpse(iris)
colnames(iris)
dim(iris)
# Create an empty vector to store means
mean_iris <- NULL
# Loop through columns of the dataset
for(i in seq_along(iris[,1])) {
  mean_iris <- c(mean_iris,mean(as.numeric(iris[i, 1:3])))
}
# Print the first 8 elements of myve
print(mean_iris[1:8])
```

Question:

1.  What is the differences between `str()` and `glimpse()`?

2.  And what is another function similar to them?

## Nested `for` loops

```{r}
n <- 5  # Define the size of the multiplication table (5x5)

# Outer loop for rows 
for (i in 1:n) {#iterates over each row index (i) from 1 to n.
  # Inner loop for columns
  for (j in 1:n) {#Inside the outer loop, iterates over each column index (j) from 1 to n.
    # Calculate product and print in a formatted way
    cat(sprintf("%2d ", i * j))
  }
  cat("\n")  # Move to the next line after each row

}
```

```{r}
# Example: Nested for loops to create a multiplication table and save as a matrix
n <- 5  # Define the size of the multiplication table (5x5)

# Initialize an empty matrix to store results
result_matrix <- matrix(0, nrow = n, ncol = n)

# Nested loops to calculate and store results
for (i in 1:n) {
  for (j in 1:n) {
    # Calculate product and store in matrix
    result_matrix[i, j] <- i * j
  }
}
```

What is the primary difference between the two nested for loop examples.
