---
title: "Script_1"
author: "Hamed Abdollahi"
date: "2024-07-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Welcome to the DAiR Workshop

The simplest task you can perform with R"

```{r}
getwd()
?setwd
```

## Package

To download and install packages from various repositories:

```{r}
# install.packages("tidyverse")
# install.packages("BiocManager")
# BiocManager::install("DESeq2")
# install.packages("devtools")
# devtools::install_github("keberwein/mlbgameday")
```

To display the paths where R looks for installed packages:

```{r}
# .libPaths()
```

To display information about the first 10 packages installed in R:

```{r}
# installed.packages()[1:3]
# head(installed.packages(), 3)
```

Displays the search path for loaded namespaces (packages) in your current R session:

```{r}
search()
```

```{r}
library()#To load and attach packages for use in R 
detach()#To detach (unload) packages
```

To uninstall (remove) packages from your R system:

```{r}
#remove.packages()#remove packages from the current library
#uninstall.packages()#remove packages from the system.
```

## Assign a value to a variable

We can store values in variables using the assignment operators `<-` and `=`, with `<-` being more prevalent in R coding conventions:

```{r}
x <- 10
y <- "Hello, world!"
z <- c(1, 2, 3)
x
y
z
```

reassign `z`

```{r}
z <- log(x)
z #Check the environment pane and look for `z` value.
```

-   variable names must follow specific rules:

    -   They cannot start with a number.

    <!-- -->

    -   They cannot contain spaces or special characters except for dot `.` and underscore `_` which is called "snake_case" style.
    - Certain words are reserved and cannot be used as variable names, such as TRUE, FALSE, NULL, if, else, function, etc. 
    + "Question for you: What are the reserved words in R?"

    -   R is case-sensitive, meaning `VariableName` and `variablename` would be considered different variables.
    
    - You can override the usual rules for variable naming in R by enclosing a name within backticks (`).
    - You should not use single or double quotes (',").
    + Question for you: Why should we not use single or double quotes (' or ")?
    
```{r}
# 1var <- 10     #Error: unexpected symbol in "1var"
#if <- "condition"     #Error: unexpected assignment in "if <-"
```
    
    
```{r}
`1var` <- 10
`if` <- "condition"
```
    

We can perform operations such as multiplication and addition:

```{r}
x*5
(x+10)/2
x <- x+1
x
```

The order of operations in R:

```{r}
3 + 5 * 2  # This evaluates to 13 because multiplication is performed before addition
(3 + 5) * 2  # This evaluates to 16 because the parentheses change the order of operations
```

R handles nested operations by following the order of operations:

exponents, multiplication, division, addition, and subtraction.

```{r}
(3 + (5 * (2 ^ 2)))
3 + 5 * 2 ^ 2   
3 + 5 * (2 ^ 2) 
```

Small decimal values

```{r}
2 / 10000 #scientific notation: 2e-4 is shorthand for 2 * 10^(-4)
```

## **Mathematical functions**

Use scientific notation for larger numbers

```{r}
6e8
```

Trigonometric functions like sine:

```{r}
sin(1)
cos(-90)
```

The natural logarithm (base e):

```{r}
log(20)
```

## Relational operators

```{r}
3 == 5  # Evaluates to FALSE

# Not equal to
3 != 5  # Evaluates to TRUE

# Greater than
3 > 5  # Evaluates to FALSE

# Less than or equal to
3 <= 5  # Evaluates to TRUE
```

-   It's generally recommended to use `==` for comparing integers:

    ```{r}
    0.1 + 0.2
    ```

    refer to [here](https://floating-point-gui.de/) the unexpected result is:

    ```         
    0.30000000000000004
    ```

    ```{r}

    # Comparing within a tolerance range
    x <- 0.1 + 0.2
    y <- 0.3
    x == y 
    tolerance <- 1e-8  # Example tolerance level
    abs(x - y) < tolerance  # Evaluates to TRUE
    ```

    Computers can represent decimal numbers with limited precision, so two numbers that appear identical when printed by R may have slightly different underlying representations due to this precision.

## Vector

One important aspect to understand in R is its vectorized nature, where variables and functions can operate on entire vectors of values.
For example:

```{r}
x1 <- c(1, 2, 3, 4, 5)  # Creating a numeric vector
x2 <- 1:5               # A numeric vector containing the values 1, 2, 3, 4, and 5
x3 <- 2^x1
y1 <- x * 2              # Multiplying each element of x by 2
y2 <- 2^(1:5)#element is 2 raised to the power of each element in the sequence from 1 to 5
z <- sqrt(x)            # Taking the square root of each element of x
x1
x2
x3
y1
y2
z
```

## **Managing Environment**

To list the names of objects (variables, functions, etc.) that are currently stored in your workspace:

```{r}
ls()
ls
```

To remove objects (variables, functions, etc.) from the workspace:

```{r}
rm(x2)
ls()
rm(list = ls())        #removes all objects from the current workspace
ls()
```

## 
